---
title: "Orange Trees and Land Temperature Data"
author: "Georges Michel, Jack Nelson, Nicole Shaver, Nathan Tuttle"
date: "April 11, 2017"
output: 
      github_document 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,clean=FALSE)
#set the working directory-
knitr::opts_knit$set(root.dir = "..//", output.dir="paper")
#ensure all required packages are installed, and if not, install them
list.of.packages <- c("plyr","ggplot2","dplyr","gdata")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages, repos="http://cran.rstudio.com/")
library(plyr)
library(dplyr)
library(ggplot2)
library(gdata)
```

## Introduction

In this study, we will examine questions related to 


## Code used to download, tidy and merge the data
```{r download_tidy_merge}
dir<-getwd()
#download the TEMP temp file and assign it to dataset name TEMPSubset
source(paste(dir,"/source/download_2.R",sep=""))
#cleanup the data-turn the dates from strings to dates
source(paste(dir,"/source/cleanup_2.R",sep=""))
#source(paste(dir,"/source/merge_ED_GDP.R",sep=""))
```

## Analysis Results

Q3ii.a. Using the data from 1900 to 2013 for average land temperatures only in the United States, we can convert the temperatures from degrees C to degrees F using the formula Temp (deg F) = Temp (deg C)* 1.8 +32.
```{r Q3iia }
#The following code chunk will keep only the data since 1900, subset only the data from the United States, rename the temperature column in the existing dataset to be explicitly in degrees C. It will then add a column with the temperature data converted to degrees C and display the first few rows of the resulting table.


Temp_clean1900 <- Temp_clean[which(Temp_clean$Date >= "1900-01-01"), ]
UStemp<-Temp_clean1900[grep("United States", Temp_clean1900$Country), ]
UStemp<-dplyr::rename(UStemp, Monthly.AverageTemp.degC=Monthly.AverageTemp)
UStemp$Monthly.AverageTemp.degF<-(UStemp$Monthly.AverageTemp.degC*1.8)+32
head(UStemp)
```
Q3ii.b. Calculating the average land temperature by Year and plotting it, we see a general trend up in average land temperatures in the United States between 1900 and 2013. 

```{r Q3iib }
#The following code chunk will take the year portion of each month, summarize the temperature data in degrees F by year, and then plot the Temperature data by year.

UStemp$Year<-as.POSIXlt(UStemp$Date)$year+1900
UStempYear<-ddply(UStemp,~Year,summarise,Temp.mean.degF=mean(Monthly.AverageTemp.degF))
ggplot(UStempYear, aes(x = Year, y=Temp.mean.degF))+geom_point()+geom_smooth()+labs(y="Average Land Temp (deg F)")+labs(title = "Average Land Temperature by Year in United States")+geom_line()+scale_x_continuous(breaks=seq(1900,2015,5))+theme(axis.text.x = element_text(size=10, angle=90))+scale_y_continuous(breaks=seq(40,56,1))
```


Q3ii.c.  Calculating the one year difference of average land temperature by year, we find that the maximum absolute temperature difference is 2.5401 degrees F, and occurred between 1920 and 1921.
```{r Q3iic }
#The code chunk below will first sort the data, by year, in ascending order. Then, it will calculate the 1 year deltas (ie 1991-1990,1992-1991), take the absolute value, and then report the max value and the years and the corresponding 2 years. 
UStempYear_sorta<-arrange(UStempYear,Year)
UStempYear_sorta$UStempYear.delta.degF <- abs(UStempYear_sorta$Temp.mean.degF -lag(UStempYear_sorta$Temp.mean.degF))
UStempYear_sorta$Year_lab<-paste(lag(UStempYear_sorta$Year),"-",UStempYear_sorta$Year)
UStempYear_sorta[which(UStempYear_sorta$UStempYear.delta == max(UStempYear_sorta$UStempYear.delta, na.rm=TRUE)), ]
```


## Conclusion

