---
title: "Orange Trees and Land Temperature Data"
author: "Georges Michel, Jack Nelson, Nicole Shaver, Nathan Tuttle"
date: "April 13, 2017"
output: 
      github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,clean=FALSE)
#set the working directory-
knitr::opts_knit$set(root.dir = "..//", output.dir="paper")
#ensure all required packages are installed, and if not, install them
list.of.packages <- c("plyr","ggplot2","dplyr","gdata","stats","reshape2")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages, repos="http://cran.rstudio.com/")
library(plyr)
library(dplyr)
library(ggplot2)
library(gdata)
library(stats)
library(reshape2)
library(graphics)
```

### Introduction
In this study, using Orange tree data from R, we will investigate orange tree growth and the relationship between tree age and circumference by tree size. We will examine land temperature differences from data that contains the monthly average temperatures since 1900 by country and by city. For the top 20 countries and cities, ranked by temperature difference, we will then examine the relationship between their minimum and maximum monthly average temperatures. Finally, we will examine land temperatures by month in the United States and look for trends in the data, and will look for temperature differences by year.



#### Code used to download and tidy data

```{r download_tidy_merge}
dir<-getwd()
#download the TEMP temp file and assign it to dataset name Temp_clean
source(paste(dir,"/source/download_2.R",sep=""))
#cleanup the data-turn the dates from strings to dates
source(paste(dir,"/source/cleanup_2.R",sep=""))
source(paste(dir,"/source/CityTemp_cleanup.R",sep=""))
```

### Orange Tree Data
```{r Orange, echo=FALSE}

summary(Orange)


Circum <- Orange$circumference

Orange_Mean <- mean(Circum)

Orange_Median <- median(Circum)

Orange_Trees <- Orange$Tree
```

The mean circumference of the Orange trees is 115.8571 centimeters trunk circumference.

```{r Mean, echo=TRUE}
Orange_Mean
```

The median circumference of the Orange trees is 115 centimeters.

```{r Median, echo=TRUE}
Orange_Median

```


## Orange Tree Circumference against Age

Here we show as the tree gets older it gets wider but none gets bigger then Tree 5 and none is smaller then Tree 1.
```{r pressure, echo=FALSE}

# Create a scatter plot circumference against age using the number of tree to define the symbol 

plot(Orange$circumference~Orange$age,
     pch=c(1,2,3,4,5)[Orange$Tree],
     col=c("blue","red","green","maroon","purple")[Orange$Tree], xlab="Age",ylab="Circumference")

# Create the legend for the trees.  
legend("topleft",c("Tree 1", "Tree 2", "Tree 3", "Tree 4", "Tree 5"),
       pch=c(1,2,3,4,5),col=c("blue","red","green","maroon","purple"))

# Create a box plot of tree circumference for each of the category of trees
boxplot(Orange$circumference~Orange$Tree,data=Orange, main="Orange Tree Circumference against Tree", 
        xlab="Tree", ylab="Circumference", pch=Orange$Tree,col=c("blue","red","green","maroon","purple"))

```
In the preceding 2 plots we see that Tree 3 is the smallest and Tree 4 is the largest.


### 3i. Find the difference between the maximum and the minimum monthly average temperatures for each country and report/visualize top 20 countries with the maximum differences for the period since 1900.

#### The following code chunk takes the Temp_clean data frame and returns T20MinMaxTemp, the top 20 countries with the greatest temperature difference between their minimum and maximum Monthly.AvgTemp observations since 1/1/1900.

##### The variables contained in T20MinMaxTemp are as follows:
* Country
* Min.Monthly.AvgTemp
* Min.Date
* Max.Monthly.AvgTemp
* Max.Date
* TempDiff

##### The steps taken to obtain T20MinMaxTemp are as follows:
* Subset Temp_clean to take only observations since 1/1/1900
* Aggregate Temp_clean1900 to receive minimum Monthly.AverageTemp
* Merge the minimum Monthly.AverageTemp value back with its Temp_clean1900 observation
* Repeat the previous two steps for maximum Monthly.AverageTemp
* Merge minimum and maximum subsets into single data frame
* Take only needed vectors for T20MinMaxTemp and rename vectors
* Calculate TempDiff
* Reorder data frame in descending order by TempDiff
* Subset to take only first 20 observations

```{r T20df}
# taking observations since 1900
Temp_clean1900 <- Temp_clean[which(Temp_clean$Date >= "1900-01-01"), ]

# arrgregate to receive min Monthly.AverageTemp by Country
Temp.agg.min <- aggregate(Monthly.AverageTemp ~ Country, Temp_clean1900, min)
# merge back with original Temp.agg.min dataframe to receive date of min observation
Temp.agg.min2 <- merge(Temp.agg.min, Temp_clean1900, by = c('Country', 'Monthly.AverageTemp'))

# arrgregate to receive max Monthly.AverageTemp by Country
Temp.agg.max <- aggregate(Monthly.AverageTemp ~ Country, Temp_clean1900, max)
# merge back with original Temp.agg.max dataframe to receive date of max observation
Temp.agg.max2 <- merge(Temp.agg.max, Temp_clean1900, by = c('Country', 'Monthly.AverageTemp'))

# merge both Temp.agg.min2 and Temp.agg.max2 together
MinMaxTemp <- merge(Temp.agg.min2, Temp.agg.max2, by = 'Country')

# remove unneeded variables and rename remaining variables in MinMaxTemp
MinMaxTemp <- MinMaxTemp[, c(1,2,4,5,7)]
names(MinMaxTemp) <- c('Country','Min.Monthly.AvgTemp','Min.Date', 'Max.Monthly.AvgTemp', 'Max.Date')

# generate difference between the max and min Monthly.AvgTemp for each observation
MinMaxTemp$TempDiff <- MinMaxTemp$Max.Monthly.AvgTemp - MinMaxTemp$Min.Monthly.AvgTemp

# order dataframe in descending order by TempDiff
MinMaxTemp <- MinMaxTemp[order(-MinMaxTemp$TempDiff), ]

# take the top 20 countries with the highest TempDiff
T20MinMaxTemp <- MinMaxTemp[1:20, ]
```

#### The following code chunk takes the T20MinMaxTemp data frame and returns T20MinMaxTempLong, the long format of T20MinMaxTemp needed for plotting.

##### The variables contained in T20MinMaxTempLong are as follows:
* Country
* Date
* TempDiff
* Min.Max
* Monthly.AvgTemp

##### The steps taken to obtain T20MinMaxTempLong are as follows:
* Converting Monthly.AvgTemp into long format for only minimum values
* Taking only the first 3 characters needed for the Min.Max indicator variable
* Repeat the previous two steps for maximum Monthly.AvgTemp values
* Renaming Min.Date and Max.Date in subsets to a common variable name Date
* Unioning both subsets together to get one long formatted data frame
* Reorder T20MinMaxTempLong to group values by Country ranked by Date
* Transform Country order to be ranked by TempDiff for plotting

```{r T20dflong}
# putting minimum metrics in long format from T20MinMaxTemp
T20MinMaxTemp1 <- melt(T20MinMaxTemp[, c(1:3,6)], id = c("Country","Min.Date", "TempDiff"), variable.name = "Min.Max", value.name = 'Monthly.AvgTemp')
# taking only first 3 character to indicate min observation
T20MinMaxTemp1$Min.Max <- substr(T20MinMaxTemp1$Min.Max, 0, 3)

# putting maximum metrics in long format from T20MinMaxTemp
T20MinMaxTemp2 <- melt(T20MinMaxTemp[, c(1,4,5,6)], id = c("Country","Max.Date", "TempDiff"), variable.name = "Min.Max", value.name = 'Monthly.AvgTemp')
# taking only first 3 characters to indicate max observation
T20MinMaxTemp2$Min.Max <- substr(T20MinMaxTemp2$Min.Max, 0, 3)

# renaming date variable in both subsets to match eachother
names(T20MinMaxTemp1)[names(T20MinMaxTemp1) == 'Min.Date'] <- 'Date'
names(T20MinMaxTemp2)[names(T20MinMaxTemp2) == 'Max.Date'] <- 'Date'

# union 2 subsets together for one long formatted T20MinMaxTemp dataframe
T20MinMaxTempLong <- union(T20MinMaxTemp1, T20MinMaxTemp2)

# reorder to group by Country and put two dates in order for plotting
T20MinMaxTempLong <- T20MinMaxTempLong[order(T20MinMaxTempLong$TempDiff,T20MinMaxTempLong$Date), ]

# reorder Country in descending order according to TempDiff
T20MinMaxTempLong <- transform(T20MinMaxTempLong, Country=reorder(Country, -TempDiff) )
```

#### The following code chunk plots the T20MinMaxTemp data frame with Country in the x axis ranked by TempDiff and Monthly.AvgTemp in the y axis.  The Min and Max values are plotting with labels indicating the year the observation occurred.  An arrow then indicates which observation occurred first (Min prior to Max = Up; Max prior to Min = Down) 

```{r T20plot}
ggplot(T20MinMaxTempLong, aes(x=Country, y=Monthly.AvgTemp, 
    group=Country, factor(TempDiff))) + 
  geom_point(size=2)+geom_path(arrow=arrow()) +
  labs(y = 'Monthly Average Temperature (Celsius)',
      title = 'Minimum & Maximum Monthly Average Temperature by Country 
      (Since 1900)') +
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5),
        plot.title = element_text(size=10, face='bold', hjust=0.5)) + 
  geom_text(data = filter_(T20MinMaxTempLong, ~Min.Max == 'Min'), 
            aes(label=format(Date,'%Y'), hjust=0.5, vjust=1.5), size = 3) +
  geom_text(data = filter_(T20MinMaxTempLong, ~Min.Max == 'Max'), 
            aes(label=format(Date,'%Y'), hjust=0.5, vjust=-0.5), size =3)
```

#### The following code chunk plots the T20MinMaxTemp data frame with Year in the x axis and Monthly.AvgTemp in the y axis.  The Min and Max Monthly.AvgTemp values are plotted by color.


```{r T20scatterplot}
ggplot(T20MinMaxTempLong, aes(x=Date, y=Monthly.AvgTemp, color=Min.Max)) + 
  geom_point(size=2) +
  labs(x = 'Year', y = 'Monthly Avgerage Temperature (Celsius)', 
      title = 'Minimum & Maximum Monthly Average Temperature vs. Year (Since 1900)',
      color = "") +
  theme(plot.title = element_text(size=12, face='bold', hjust=0.5))
```

### Q3ii.a. Select a subset of data called "UStemp" where US land temperatures from 01/01/1990 in Temp data. Use UStemp dataset to answer the followings. Create a new column to display the monthly average land temperatures in Fahrenheit (°F).

#### Using the data from 1900 to 2013 for average land temperatures only in the United States, we can convert the temperatures from degrees C to degrees F using the formula Temp (deg F) = Temp (deg C)* 1.8 +32.

```{r Q3iia}

#The following code chunk will keep only the data since 1900, subset only the data from the United States, rename the temperature column in the existing dataset to be explicitly in degrees C. It will then add a column with the temperature data converted to degrees C and display the first few rows of the resulting table.


UStemp<-Temp_clean1900[grep("United States", Temp_clean1900$Country), ]
UStemp<-dplyr::rename(UStemp, Monthly.AverageTemp.degC=Monthly.AverageTemp)
UStemp$Monthly.AverageTemp.degF<-(UStemp$Monthly.AverageTemp.degC*1.8)+32
head(UStemp)
```

### Q3ii.b. Calculate average land temperature by year and plot it. The original file has the average land temperature by month.

#### Calculating the average land temperature by Year and plotting it, we see a general trend up in average land temperatures in the United States between 1900 and 2013. 

```{r Q3iib }
#The following code chunk will take the year portion of each month, summarize the temperature data in degrees F by year, and then plot the Temperature data by year.

UStemp$Year<-as.POSIXlt(UStemp$Date)$year+1900
UStempYear<-ddply(UStemp,~Year,summarise,Temp.mean.degF=mean(Monthly.AverageTemp.degF))
ggplot(UStempYear, aes(x = Year, y=Temp.mean.degF))+
  geom_point()+
  geom_smooth()+
  labs(y="Average Land Temp (deg F)")+
  labs(title = "Average Land Temperature by Year in United States")+
  geom_line()+scale_x_continuous(breaks=seq(1900,2015,5))+
  theme(axis.text.x = element_text(size=10, angle=90))+
  scale_y_continuous(breaks=seq(40,56,1))
```

### Q3ii.c. Calculate the one year difference of average land temperature by year and provide the maximum difference (value) with corresponding two years.

#### Calculating the one year difference of average land temperature by year, we find that the maximum absolute temperature difference is 2.5401 degrees F, and occurred between 1920 and 1921.

```{r Q3iic }
#The code chunk below will first sort the data, by year, in ascending order. Then, it will calculate the 1 year deltas (ie 1991-1990,1992-1991), take the absolute value, and then report the max value and the years and the corresponding 2 years.


UStempYear_sorta<-arrange(UStempYear,Year)
UStempYear_sorta$UStempYear.delta.degF <- abs(UStempYear_sorta$Temp.mean.degF -lag(UStempYear_sorta$Temp.mean.degF))
UStempYear_sorta$Year_lab<-paste(lag(UStempYear_sorta$Year),"-",UStempYear_sorta$Year)
UStempYear_sorta[which(UStempYear_sorta$UStempYear.delta == max(UStempYear_sorta$UStempYear.delta, na.rm=TRUE)), ]
```

### 3iii. Download CityTemp data set at box.com. Find the difference between the maximum and the minimum temperatures for each major city and report/visualize top 20 cities with maximum differences for the period since 1900.

#### The following code chunk takes the CityTemp_clean data frame and returns T20MinMaxCityTemp, the top 20 cities with the greatest temperature difference between their minimum and maximum Monthly.AvgTemp observations since 1/1/1900.

##### The variables contained in T20MinMaxCityTemp are as follows:
* City
* Min.Monthly.AvgTemp
* Country
* Latitude
* Min.Date
* Max.Monthly.AvgTemp
* Max.Date
* TempDiff

##### The steps taken to obtain T20MinMaxCityTemp are as follows:
* Subset CityTemp_clean to take only observations since 1/1/1900
* Aggregate CityTemp_clean1900 to receive minimum Monthly.AverageTemp
* Merge the minimum Monthly.AverageTemp value back with its CityTemp_clean1900 observation
* Repeat the previous two steps for maximum Monthly.AverageTemp
* Merge minimum and maximum subsets into single data frame
* Take only needed vectors for T20MinMaxCityTemp and rename vectors
* Calculate TempDiff
* Reorder data frame in descending order by TempDiff
* Subset to take only first 20 observations

```{r T20Citydf}
# taking observations since 1900
CityTemp_clean1900 <- CityTemp_clean[which(CityTemp_clean$Date >= "1900-01-01"), ]

# arrgregate to receive min Monthly.AverageTemp by City
CityTemp.agg.min <- aggregate(Monthly.AverageTemp ~ City, CityTemp_clean1900, min)
# merge back with original CityTemp.agg.min dataframe to receive date of min observation
CityTemp.agg.min2 <- merge(CityTemp.agg.min, CityTemp_clean1900, by = c('City', 'Monthly.AverageTemp'))

# arrgregate to receive max Monthly.AverageTemp by City
CityTemp.agg.max <- aggregate(Monthly.AverageTemp ~ City, CityTemp_clean1900, max)
# merge back with original Temp.agg.max dataframe to receive date of max observation
CityTemp.agg.max2 <- merge(CityTemp.agg.max, CityTemp_clean1900, by = c('City', 'Monthly.AverageTemp'))

# merge both Temp.agg.min2 and Temp.agg.max2 together
MinMaxCityTemp <- merge(CityTemp.agg.min2, CityTemp.agg.max2, by = 'City')

# remove unneeded variables and rename remaining variables in MinMaxTemp
MinMaxCityTemp <- MinMaxCityTemp[, c(1,2,4,5,7,8,13)]
names(MinMaxCityTemp) <- c('City', 'Min.Monthly.AvgTemp', 'Country', 'Latitude', 
                       'Min.Date', 'Max.Monthly.AvgTemp', 'Max.Date')

# generate difference between the max and min Monthly.AvgTemp for each observation
MinMaxCityTemp$TempDiff <- MinMaxCityTemp$Max.Monthly.AvgTemp - MinMaxCityTemp$Min.Monthly.AvgTemp

# order dataframe in descending order by TempDiff
MinMaxCityTemp <- MinMaxCityTemp[order(-MinMaxCityTemp$TempDiff), ]

# take the top 20 countries with the highest TempDiff
T20MinMaxCityTemp <- MinMaxCityTemp[1:20, ]
```

#### The following code chunk takes the T20MinMaxCityTemp data frame and returns T20MinMaxCityTempLong, the long format of T20MinMaxCityTemp needed for plotting.

##### The variables contained in T20MinMaxCityTempLong are as follows:
* City
* Country
* Latitude
* Date
* TempDiff
* Min.Max
* Monthly.AvgTemp

##### The steps taken to obtain T20MinMaxCityTempLong are as follows:
* Converting Monthly.AvgTemp into long format for only minimum values
* Taking only the first 3 characters needed for the Min.Max indicator variable
* Repeat the previous two steps for maximum Monthly.AvgTemp values
* Renaming Min.Date and Max.Date in subsets to a common variable name Date
* Unioning both subsets together to get one long formatted data frame
* Reorder T20MinMaxCityTempLong to group values by City ranked by Date
* Transform City order to be ranked by TempDiff for plotting

```{r T20Citydflong}
# putting minimum metrics in long format from T20MinMaxCityTemp
T20MinMaxCityTemp1 <- melt(T20MinMaxCityTemp[, c(1:5,8)], id = 
                           c("City", "Country", "Latitude", "Min.Date", "TempDiff"), 
                           variable.name = "Min.Max", value.name = 'Monthly.AvgTemp')
# taking only first 3 character to indicate min observation
T20MinMaxCityTemp1$Min.Max <- substr(T20MinMaxCityTemp1$Min.Max, 0, 3)

# putting maximum metrics in long format from T20MinMaxCityTemp
T20MinMaxCityTemp2 <- melt(T20MinMaxCityTemp[, c(1,3,4,6:8)], id = 
                           c("City", "Country", "Latitude", "Max.Date", "TempDiff"), 
                           variable.name = "Min.Max", value.name = 'Monthly.AvgTemp')
# taking only first 3 characters to indicate max observation
T20MinMaxCityTemp2$Min.Max <- substr(T20MinMaxCityTemp2$Min.Max, 0, 3)

# renaming date variable in both subsets to match eachother
names(T20MinMaxCityTemp1)[names(T20MinMaxCityTemp1) == 'Min.Date'] <- 'Date'
names(T20MinMaxCityTemp2)[names(T20MinMaxCityTemp2) == 'Max.Date'] <- 'Date'

# union 2 subsets together for one long formatted T20MinMaxCityTemp dataframe
T20MinMaxCityTempLong <- union(T20MinMaxCityTemp1, T20MinMaxCityTemp2)

# reorder to group by City and put two dates in order for plotting
T20MinMaxCityTempLong <- T20MinMaxCityTempLong[order(-T20MinMaxCityTempLong$TempDiff,
                                               T20MinMaxCityTempLong$Date), ]

# reorder City in descending order according to TempDiff
T20MinMaxCityTempLong <- transform(T20MinMaxCityTempLong, City=reorder(City, -TempDiff) )
```

#### The following code chunk plots the T20MinMaxCityTemp data frame with City in the x axis ranked by TempDiff and Monthly.AvgTemp in the y axis.  The Min and Max values are plotting with labels indicating the year the observation occurred.  An arrow then indicates which observation occurred first (Min prior to Max = Up; Max prior to Min = Down). 

```{r T20Cityplot}
ggplot(T20MinMaxCityTempLong, aes(x=City, y=Monthly.AvgTemp, 
                               group=City, factor(TempDiff))) + 
  geom_point(size=2)+geom_path(arrow=arrow()) +
  labs(y = 'Monthly Average Temperature (Celsius)',
       title = 'Minimum & Maximum Monthly Average Temperature by City (Since 1900)') +
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0.5),
        plot.title = element_text(size=10, face='bold', hjust=0.5)) + 
  geom_text(data = filter_(T20MinMaxCityTempLong, ~Min.Max == 'Min'), 
            aes(label=format(Date,'%Y'), hjust=0.5, vjust=1.5), size = 3) +
  geom_text(data = filter_(T20MinMaxCityTempLong, ~Min.Max == 'Max'), 
            aes(label=format(Date,'%Y'), hjust=0.5, vjust=-0.5), size =3)
```

#### The following code chunk plots the T20MinMaxCityTemp data frame with Year in the x axis and Monthly.AvgTemp in the y axis.  The Min and Max Monthly Average Temperature (Monthly.AvgTemp) values are plotted by color.


```{r T20Cityscatterplot}
ggplot(T20MinMaxCityTempLong, aes(x=Date, y=Monthly.AvgTemp, color=Min.Max)) + 
  geom_point(size=2) +
  labs(x = 'Year', y = 'Monthly Avgerage Temperature (Celsius)', 
       title = 'Minimum & Maximum Monthly Average Temperature vs. Year (Since 1900)',
       color = "") +
  theme(plot.title = element_text(size=12, face='bold', hjust=0.5))
```

### 3iv. Compare the two graphs in (i) and (iii) and comment it.

#### The initial plots have arrows indicating a minimum or maximum Monthly Average Temperature (Monthly.AvgTemp).  The arrows in the plot show commonalities between the top 20 countries and top 20 cities with the highest TempDiff since 1/1/1900.  Eighteen out of 20 countries and 15 out of 20 cities have their maximum Monthly Average Temperature (Monthly.AvgTemp) occurring later than the minimum Monthly Average Temperature (Monthly.AvgTemp).  The scatter plot showing minimum and maximum Monthly Average Temperature (Monthly.AvgTemp) by year is even more telling. There appears to be a grouping for maximum values between 1980-present (18 out of 20 countries, 16 out of 20 cities).  Without the use of advanced statistics and only using visual inspection, it appears that the Monthly Average Temperature (Monthly.AvgTemp) of the top 20 countries and cities with the highest TempDiff since 1/1/1900 has been rising in the last 40 years.

## Conclusion
In this study, we find visual evidence that further investigation is warranted to determine if monthly average temperatures have been rising since January 1900. This assertion is based on data from the top 20 countries and cities with the greatest temperature differences. Additionally, we we find that in the United States, the maximum absolute temperature difference by year is 2.5401 degrees F (1920-1921), and that there is a general trend up in average land temperature.